Polynomial phase estimation by phase unwrapping 2: Asymptotic normality and simulations

### 2012-10-29 paper.tex
minor edits

The requirement that $\Phi_1, \dots, \Phi_N$ be identically distributed will typically hold only when the complex random variables $X_1, \dots, X_N$ are identically distributed and circularly symmetric.  It would be possible to drop the assumption that $\Phi_1, \dots, \Phi_N$ be identically distributed, but this complicates the theorem statement and the proof.  In the interest of simplicity we only consider the case when $\Phi_1, \dots, \Phi_N$ are identically distributed here.  If $X_n$ is circularly symmetric with density function nonincreasing with magnitude $\abs{X_n}$, then, the corresponding $\Phi_n$ necessarily has zero intrinsic mean~\cite[Theorem 5.2, page 78]{McKilliam2010thesis}.  Thus, our theorem covers commonly used distributions for $X_1, \dots, X_N$, such as the normal distribution.

>>>>

The requirement that $\Phi_1, \dots, \Phi_N$ be identically distributed will typically hold only when the complex random variables $X_1, \dots, X_N$ are identically distributed and circularly symmetric.  It would be possible to drop the assumption that $\Phi_1, \dots, \Phi_N$ be identically distributed, but this complicates the theorem statement and the proof.  In the interest of simplicity we only consider the case when $\Phi_1, \dots, \Phi_N$ are identically distributed here.  If $X_n$ is circularly symmetric with density function nonincreasing with magnitude $\abs{X_n}$, then, the corresponding $\Phi_n$ necessarily has zero intrinsic mean~\cite[Theorem~5.2,~page~78]{McKilliam2010thesis}.  Thus, our theorem covers commonly used distributions for $X_1, \dots, X_N$, such as the normal distribution.

Although we will not prove it here the assumption that $\Phi_1,\dots,\Phi_N$ have zero intrinsic mean is not only sufficient, but also necessary, for if $\Phi_1,\dots,\Phi_N$ have intrinsic mean $x \in [-\nicefrac{1}{2},\nicefrac{1}{2})$ with $x \neq 0$ then $\fracpart{\widehat{\lambda}_{0,N} - x}\rightarrow 0$ almost surely as $N\rightarrow\infty$, and so $\widehat{\lambda}_{0,N}$ does not converge to zero.  On the other hand if $\Phi_1,\dots,\Phi_N$ do not have an intrinsic mean then $\widehat{\lambda}_{0,N}$ will not converge.

### 2012-10-29 paper2.tex
minor edits

We now describe the least squares unwrapping (LSU) estimator of the polynomial coefficients. Recall that we desire to estimate the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from the noisy samples $Y_1, \dots, Y_N$ given in~\eqref{eq:Y_nsamplednoisey}.  We take the complex argument of the $Y_n$ and divide by $2\pi$ to obtain
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) }
\end{equation}
where $\angle{z}$ denotes the complex argument of the complex number $z$, and 
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n)
\] 
are random variables representing the phase noise induced by the $X_n$~\cite{Tretter1985,Quinn2009_dasp_phase_only_information_loss}.  If the distribution of $X_n$ is circularly symmetric (i.e., the angle $\angle X_n$ is uniformly distributed on $[-\pi, \pi)$ and is independent of the magnitude $\abs{X_n}$) then the distribution of $\Phi_n$ is the same as the distribution of $\tfrac{1}{2\pi}\angle(1 + \rho^{-1}X_n)$.  If the $X_1, \dots, X_N$ are circularly symmetric and identically distributed, then $\Phi_1, \dots, \Phi_n$ are also identically distributed.

Let $\mubf$ be the vector $[\mu_0, \mu_1, \dots, \mu_m]$ and put,
\begin{equation} \label{eq:sumofsquaresfunction}
SS(\mubf) = \sum_{n=1}^{N}\fracpart{  \Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }^{2}.
\end{equation}
The least squares unwrapping estimator is defined as those coefficients $\widehat{\mu}_0, \dots, \widehat{\mu}_m$ that minimise $SS$ over the identifiable region $B$. That is, the LSU estimator is,
\begin{equation}\label{eq:hatmubfLSUdefn}
\widehat{\mubf} = \arg \min_{\mubf \in B} SS(\mubf). 
\end{equation}

It is shown in~\cite[Sec~8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} how this minimisation problem can be posed as that of computing a nearest lattice point in a particular lattice. Polynomial time algorithms that compute the nearest point are described in~\cite[Sec.~4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.  In this paper we focus on the asymptotic statistical properties of the LSU estimator, rather than computational aspects.

The next theorem describes the asymptotic properties of this estimator.  Before we give the proof it is necessary to understand some of the properties of the phase noise $\Phi_1,\dots,\Phi_N$, which are \emph{circular} random variables with support on $[-\nicefrac{1}{2}, \nicefrac{1}{2})$~\cite{McKilliam2010thesis,McKilliam_mean_dir_est_sq_arc_length2010,Mardia_directional_statistics,Fisher1993}.  Circular random variables are often considered modulo $2\pi$ and therefore have support $[-\pi, \pi)$ with $-\pi$ and $\pi$ being identified as equivalent.  Here we instead consider circular random variables modulo 1 with support $[-\nicefrac{1}{2}, \nicefrac{1}{2})$ and with $-\nicefrac{1}{2}$ and $\nicefrac{1}{2}$ being equivalent.  This is nonstandard but it allows us to use notation such as $\round{\cdot}$ for rounding and $\fracpart{\cdot}$ for the centered fractional part in a convenient way.   %The random variable $2\pi \Phi_n$ has support on $[-\pi, \pi)$ and can be identified as a \emph{circular random variable}~\cite{Mardia_directional_statistics,Fisher1993}.  %Let $f$ be the probability density function (pdf) of $\Phi_n$.  Then $f$ takes nonzero values only on $[-\nicefrac{1}{2}, \nicefrac{1}{2})$.  It is instructive to think of $f$ as describing a describing a distribution on a circle.  For example, when the noise term $X_n$ is complex Gaussian with independent and identically distributed real and imaginary parts, the distribution of $\Phi_n$ is the projected Guassian~\cite[Section 5.6.1]{McKilliam2010thesis}\cite[page 46]{Mardia_directional_statistics}.  It is instrutive the consider a periodic version of $f$, that we write as $f(\fracpart{x})$, which has period equal to one, because the fractional part function $\fracpart{x}$ has period one.

The \emph{intrinsic mean} or \emph{Fr\'{e}chet mean} of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,bwhk07a,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
\begin{equation}\label{eq:intrmeandefn}
 \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{equation}
and the \emph{intrinsic variance} is
\[
\sigma_{\text{intr}}^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
\]
where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  We are now equipped to state the asymptotic properties of the LSU estimator.

@@

\begin{theorem} \label{thm:asymp_proof} (Asymptotic normality)
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$.  If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k = \{0, 1, \dots, m\}$.
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  The proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  %Before giving the proof we make the following remarks.
The theorem gives conditions on the \emph{dealiased} difference $\dealias(\tilde{\mubf} - \widehat{\mubf})$ between the true coefficients $\tilde{\mubf}$ and the estimated coefficients $\widehat{\mubf}$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phase $\pm 0.5$ are actually the same.  In this case 
\[
\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01.
\] 
The same reasoning holds for $m > 0$.

>>>>

As in~\cite{McKilliam_pps1_2012} the least squares unwrapping (LSU) estimator of the polynomial coefficients is the minimiser
\begin{equation}\label{eq:hatmubfLSUdefn}
\widehat{\mubf} = \arg \min_{\mubf \in B} SS(\mubf)
\end{equation}
where the objective function
\begin{equation} \label{eq:sumofsquaresfunction}
SS(\mubf) = \sum_{n=1}^{N}\fracpart{  \Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }^{2},
\end{equation}
and
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) },
\end{equation}
is the phase (divided by $2\pi$) of the observation $Y_1,\dots,Y_N$, and
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n).
\] 
are random variables representing the \emph{phase noise} observed at the reciever~\cite{Tretter1985}.  We denote by $\angle{z}$ the complex argument of the complex number $z$, and by $\fracpart{x} = x - \round{x}$ the (centered) fractional part of the real number $x$, and by $\round{x}$ the nearest integer to $x$.  We have chosen to round half integers up here.  The next theorem describes the asymptotic distribution of the LSU estimator.

%It is shown in~\cite[Sec~8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} how the minimisation problem in~\eqref{eq:hatmubfLSUdefn} can be posed as that of computing a nearest lattice point in a particular lattice. Polynomial time algorithms that compute the nearest point are described in~\cite[Sec.~4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.  In this paper we focus on the asymptotic statistical properties of the LSU estimator, rather than computational aspects.

%The next theorem describes the asymptotic properties of this estimator.  Before we give the proof it is necessary to understand some of the properties of the phase noise $\Phi_1,\dots,\Phi_N$, which are \emph{circular} random variables with support on $[-\nicefrac{1}{2}, \nicefrac{1}{2})$~\cite{McKilliam2010thesis,McKilliam_mean_dir_est_sq_arc_length2010,Mardia_directional_statistics,Fisher1993}.  Circular random variables are often considered modulo $2\pi$ and therefore have support $[-\pi, \pi)$ with $-\pi$ and $\pi$ being identified as equivalent.  Here we instead consider circular random variables modulo 1 with support $[-\nicefrac{1}{2}, \nicefrac{1}{2})$ and with $-\nicefrac{1}{2}$ and $\nicefrac{1}{2}$ being equivalent.  This is nonstandard but it allows us to use notation such as $\round{\cdot}$ for rounding and $\fracpart{\cdot}$ for the centered fractional part in a convenient way.   %The random variable $2\pi \Phi_n$ has support on $[-\pi, \pi)$ and can be identified as a \emph{circular random variable}~\cite{Mardia_directional_statistics,Fisher1993}.  %Let $f$ be the probability density function (pdf) of $\Phi_n$.  Then $f$ takes nonzero values only on $[-\nicefrac{1}{2}, \nicefrac{1}{2})$.  It is instructive to think of $f$ as describing a describing a distribution on a circle.  For example, when the noise term $X_n$ is complex Gaussian with independent and identically distributed real and imaginary parts, the distribution of $\Phi_n$ is the projected Guassian~\cite[Section 5.6.1]{McKilliam2010thesis}\cite[page 46]{Mardia_directional_statistics}.  It is instrutive the consider a periodic version of $f$, that we write as $f(\fracpart{x})$, which has period equal to one, because the fractional part function $\fracpart{x}$ has period one.

% The \emph{intrinsic mean} or \emph{Fr\'{e}chet mean} of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,bwhk07a,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
% \begin{equation}\label{eq:intrmeandefn}
%  \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% % %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
% \end{equation}
% and the \emph{intrinsic variance} is
% \[
% \sigma_{\text{intr}}^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
% \]
% where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  We are now equipped to state the asymptotic properties of the LSU estimator.

@@

\begin{theorem} \label{thm:asymp_proof} (Asymptotic normality)
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero mean, variance $\sigma^2$, and pdf $f$ such that $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$.  Suppose also that  $N^k \widehat{\lambda}_{k,N} \rightarrow 0$ in probability as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.  Then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k = \{0, 1, \dots, m\}$.
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  The proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  %Before giving the proof we make the following remarks.
The theorem gives conditions on the \emph{dealiased} difference $\dealias(\tilde{\mubf} - \widehat{\mubf})$ between the true coefficients $\tilde{\mubf}$ and the estimated coefficients $\widehat{\mubf}$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phase $\pm 0.5$ are actually the same.  In this case 
\[
\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01.
\] 
The same reasoning holds for $m > 0$.
